---
title: "Chicago Housing"
author: "Annelise Escher"
date: "5/29/2022"
output:
  pdf_document: default
  html_document:
    number_sections: yes
    keep_md: yes
---
```{r message=FALSE, warning=FALSE}
rm(list = ls())
library(tidyverse)
library(stargazer)
library(readxl)
library(knitr)
library(lubridate)
library(dplyr)
library(devtools)
library(sf)
library(janitor)
library(MASS)
library(lmtest)
library(sandwich)
knitr::opts_chunk$set(echo = FALSE)
```

## Reading in, cleaning, and combining data 

Reading in data
```{r}
# WBEZ data portal 
chicago_purchases <- read_excel("Chicago Purchase Orginations 2012-18_Public File.xlsx", 
                                sheet = "Data", 
                                col_types = "text") 
# NRS
redlining_score <- read_excel("Historic Redlining Score 2010B.xlsx")
# Converting Census tracts to zipcodes for visualizations
census_to_zip <- read_excel("ZIP_TRACT_122018.xlsx") %>% mutate(GEOID10 = tract) 
```

Cleaning main datasets
```{r}
chi_redlining_score <- redlining_score %>% filter(METRO_NAME == "Chicago-Naperville-Elgin, IL-I") 

chicago_purchases <- chicago_purchases %>% 
  clean_names() %>%
  mutate(GEOID10 = census_tract) %>%
    dplyr::select(-contains('denial'),
                  -contains('applicant'), 
                  -parent_name_panel, -top_holder_name)
```

Adding redlining grade and zip codes to purchase origination dataframe.
```{r}
chicago_purchases <- chicago_purchases %>%
  left_join(redlining_score, by = "GEOID10") 
chic_purchases_w_zip <- chicago_purchases %>%
  left_join(census_to_zip, by = "GEOID10") %>%
  mutate(zip = as.character(zip)) 
```

## Data visualizations 

Creating dataframe for visualization by zip code.
```{r}
zip_stats <- chic_purchases_w_zip %>%
  mutate(loan_amount = as.numeric(loan_amount)) %>%
  group_by(zip) %>%
  summarise(loan_total = sum(loan_amount), 
            white_percent_avg = mean(as.numeric(tract_percent_population_white)), 
            black_percent_avg = mean(as.numeric(tract_percent_population_black)), 
            latino_percent_avg = mean(as.numeric(tract_percent_population_latino)), 
            asian_percent_avg = mean(as.numeric(tract_percent_population_asian))) %>%
  mutate(zip = as.character(zip), 
         majority_race = ifelse(white_percent_avg >= 0.5, "white", 
                         ifelse(black_percent_avg >= 0.5, "black", 
                         ifelse(latino_percent_avg >= 0.5, "latino", "no_majority")))) %>%
  arrange(desc(loan_total))
head(zip_stats, 3)
```

Recreating WBEZ figure. Map of Chicago zip codes color coded by the dollar amount of mortgage loans made to that area from 2012-2018. 
```{r}
zip_stats_for_map <- zip_stats %>%
  dplyr::select(zip, loan_total) %>%
  rename(region = zip, value = loan_total) %>%
  transform(region = as.numeric(region)) %>%
  mutate(region = str_trim(region)) %>%
  filter(region < 60666)

jpeg(file="zip_code.jpeg")
choroplethrZip::zip_choropleth(zip_stats_for_map,
                               zip_zoom = zip_stats_for_map$region,
                               title = "Home Mortgage Loans, 2012-2018",
                               legend = "Dollar Amounts"
)
dev.off()
```

Bar chart of total home mortgage lending to Chicago zip codes, color coding for majority race of neighborhood. 
```{r}
jpeg(file="zip_code_by_race.jpeg")
zip_stats %>%
  ggplot(aes(x = reorder(zip, loan_total), y = loan_total, fill = majority_race)) + 
  geom_bar(stat = 'identity') + 
  scale_y_continuous(breaks = seq(0, 8000000000, by = 1800000000), labels = scales::comma) + 
  coord_flip() + 
  scale_fill_manual(values=c("aquamarine3", "brown3", "darkgoldenrod2", "darkblue"))  +
  theme(panel.background = element_blank()) +
  theme(axis.text.y=element_blank(), #remove x axis labels
        axis.ticks.y=element_blank()) + 
  labs(title = "Total Mortgage Loans by Zip Code, 2012-2018", 
       x = "Chicago Zip Codes", 
       y = "Total Loans in Dollars", 
       fill = "Majority Race of Zip Code") 
dev.off()
```

Creating dataframe of modern home mortgage lending by census tract used for statistical analysis. 
```{r}
census_stats <- chicago_purchases %>%
  mutate(loan_amount = as.numeric(loan_amount)) %>%
  group_by(census_tract) %>%
  summarise(loan_total = sum(loan_amount), 
            white_percent_avg = mean(as.numeric(tract_percent_population_white)), 
            black_percent_avg = mean(as.numeric(tract_percent_population_black)), 
            latino_percent_avg = mean(as.numeric(tract_percent_population_latino)), 
            asian_percent_avg = mean(as.numeric(tract_percent_population_asian)),
            tract_pop = mean(as.numeric(tract_population)), 
            med_family_income = mean(as.numeric(tract_median_family_income)), 
            avg_loan_amount = mean(as.numeric(loan_amount))
            ) %>%
  mutate(census_tract = as.character(census_tract), 
         majority_race = ifelse(white_percent_avg >= 0.5, "white", 
                         ifelse(black_percent_avg >= 0.5, "black", 
                         ifelse(latino_percent_avg >= 0.5, "latino", "no_majority")))) %>%
  arrange(desc(loan_total)) %>%
  mutate(id = row_number())
head(census_stats, 3)
```

Creating dataframe including redlining scores by census tract for statistical analysis. 
```{r}
nrow(census_stats)
for_regression <- census_stats %>%
  left_join(chi_redlining_score, by = c("census_tract" = "GEOID10")) %>% # adding redlining scores to census tract data
  filter(!is.na(INTERVAL2010)) %>%
  rename(redline_1930_value = INTERVAL2010) %>%  
  mutate(majority_race = as.factor(majority_race), relevel(majority_race, ref = "no_majority")) %>%
  mutate(loan_total = ifelse(loan_total <= 535221000, loan_total, 535221000)) %>% # top coding two upper  loan_total values for the South Loop and Fulton Market neighborhoods 
  mutate(loans_millions = loan_total/1000000) %>%
  mutate(value_2 = ifelse(redline_1930_value <= 2, 1, 0)) %>% # creating binary variable for 1 or 2 HOLC grad
  mutate(value_3 = ifelse(redline_1930_value == 3, 1, 0)) %>% # creating binary variable for 3 HOLC grad
  mutate(value_4 = ifelse(redline_1930_value == 4, 1, 0)) # creating binary variable for 4 HOLC grad
nrow(for_regression) # number of rows after removing NAs
```

## Regression 1: Simple OLS
```{r}
for_regression$majority_race <- relevel(for_regression$majority_race, ref = "no_majority") # setting no majority racial group as the baseline category to compare findings for majority White, Black, and Latino neighborhoods 

ols_cont_reg <- lm(loans_millions ~ HRS2010 + tract_pop + med_family_income + avg_loan_amount + majority_race, data = for_regression)
summary(ols_cont_reg)
ols_categ_reg <- lm(loans_millions ~ redline_1930_value + tract_pop + med_family_income + avg_loan_amount + majority_race, data = for_regression)
summary(ols_categ_reg)
```

## Regression 2: Past category of HOLC grade as predictor for modern recreated distribution 

Creating a new variable showing modern census tracts categorized by the same distribution as the 1930s HOLC grade for chicago. 
```{r}
table(for_regression$redline_1930_value) # number of census tracts in each of the four HOLC grades
table(for_regression$redline_1930_value)/nrow(for_regression) # proportion of tracts in that HOLC grade
for_regression <- for_regression %>%
  mutate(ranking_2010 = ifelse(loan_total >= quantile(census_stats$loan_total, 
                                                      probs = 0.9945, na.rm = TRUE), 1,
                        ifelse(loan_total >= quantile(census_stats$loan_total, 
                                                      probs = 0.9056, na.rm = TRUE), 2, 
                        ifelse(loan_total >= quantile(census_stats$loan_total, 
                                                      probs = 0.3908, na.rm = TRUE), 3, 4))))
# creating new categorical variable on 1-4 grading scale for level of home mortgage lending
```

Table comparing number of census tracts in each of the four HOLC grades. This shows us that the top census tracts by level of 
```{r}
table(for_regression$ranking_2010)
ftable(xtabs(~ redline_1930_value + ranking_2010, data = for_regression))
```


```{r}
grade_categ_reg <- lm(ranking_2010 ~ value_3 + value_4, data=for_regression)
summary(grade_categ_reg)
grade_cont_reg <- lm(ranking_2010 ~ HRS2010, data=for_regression)
summary(grade_cont_reg)
```

## Regression 3: Linear Probability Model 
```{r}
cat_2 <- lm(value_2 ~ redline_1930_value, data = for_regression)
coeftest(cat_2, vcov = vcovHC(cat_2, type="HC1"))
cat_3 <- lm(value_3 ~ redline_1930_value, data = for_regression)
coeftest(cat_3, vcov = vcovHC(cat_3, type="HC1"))
cat_4 <- lm(value_4 ~ redline_1930_value, data = for_regression)
coeftest(cat_4, vcov = vcovHC(cat_4, type="HC1"))

cont_2 <- lm(value_2 ~ redline_1930_value, data = for_regression)
coeftest(cont_2, vcov = vcovHC(cont_2, type="HC1"))
cont_3 <- lm(value_3 ~ redline_1930_value, data = for_regression)
coeftest(cont_3, vcov = vcovHC(cont_3, type="HC1"))
cont_4 <- lm(value_4 ~ redline_1930_value, data = for_regression)
coeftest(cont_4, vcov = vcovHC(cont_4, type="HC1"))
```

## Regression 4: Ordered logit 
```{r}
ord_logit_reg <- polr(as.factor(ranking_2010) ~ redline_1930_value, data = for_regression, Hess = TRUE)
summary(ord_logit_reg)
```
Source: https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/

## Regression summaries in stargazer for Latex
```{r mylatextable, results = "asis"}
# Regression 1 - OLS
stargazer(ols_cont_reg, ols_categ_reg,
         type = 'latex',
         omit = 'Constant',
         header = FALSE,
         single.row = TRUE, # to put coefficients and standard errors on same line 
         no.space = TRUE, # to remove the spaces after each line of coefficients 
         column.sep.width = "3pt", # to reduce column width
         font.size = "small",
         column.labels = c("Continous Redline Grade", "Categorical Redline Grade"),
         covariate.labels = c("Continous Redline Grade", "Redline Categories", "Tract Population", "Average Family Income", "Average Loan Amount", "Majority Black", "Majority Latino", "Majority White"))

# Regression 2 - HOLC grades 
stargazer(grade_cont_reg, grade_categ_reg,
         type = 'latex',
         omit = 'Constant',
         header = FALSE,
         single.row = TRUE, # to put coefficients and standard errors on same line 
         no.space = TRUE, # to remove the spaces after each line of coefficients 
         dep.var.labels = c("Modern Lending Ranking"),
         column.sep.width = "3pt", # to reduce column width
         font.size = "small",
         covariate.labels = c("Continous Variable", "Grade 3", "Grade 4"))

# Regression 4 - ordered logit 
stargazer(ord_logit_reg,
         type = 'latex')
```

